Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-02-10 ms/ms.tex
comments

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
pathogens with higher reproduction numbers ($\rzero$) --- that is, the number of secondary infections caused by a single infected host over the course of its infectious period --- will tend to increase in prevalence relative to strains with lower reproductive ratios.
Pathogens can increase their reproduction numbers either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
transmission and disease-induced mortality rates are both driven by the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.

Fraser \etal\ have applied these ideas to HIV,
showing that the virus appears to satisfy the prerequisites of
the tradeoff theory.
The \emph{set-point viral load} (SPVL: i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection) is a measurable proxy for the rate of HIV within-host
reproduction. Higher viral loads are correlated with 
shorter progression times (higher virulence).
Studies of discordant partnerships --- stable
sexual partnerships with one infected and one uninfected partner --- %
have shown that SPVL is (1) correlated with faster transmission
(the time to infection of the uninfected partner is shorter when
their partner has a higher SPVL) and (2) heritable
(when the uninfected member of
the couple does become infected, their SPVL is similar to their
partner's).
Furthermore, the rate of transmission increases with
increasing SPVL (and hence with decreasing progression time)
but with a decreasing slope as progression time decreases,
fulfilling the requirements of the tradeoff theory
 \cite{Fraser+2007}.
Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

Our models
explicitly track the evolution of the distribution of $\log_{10}$ SPVL
(which we denote as $\alpha$)
rather than the rate of progression to
AIDS itself (hereafter we use ``virulence'' to denote \Lspvl).
We use a single-stage model of HIV that assumes constant
infectivity over the course of an exponentially
distributed infectious period. This assumption contrasts
with Shirreff \etal's previous model which
explicitly tracked three stages of HIV infection (primary,
asymptomatic, and AIDS) and used a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. 
Shirreff \etal\ allowed the transmission rate
and infection duration to depend on virus load only during the
asymptomatic stage. In order to adapt their parameterization to our
single-stage model, we used their estimates to compute the SPVL-dependent transmission
rate and duration during the asymptomatic stage and 
then derived the overall duration of the infectious period
as the sum of the three stage durations and the 
average transmission rate as the duration-weighted 
average of the three stage-specific transmission rates. 
Thus the within-couple transmission rate, $\beta$
(see ``Contact Structure'' below)
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL
(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\label{eq:hillfuns}
\end{equation}
where $V_{\alpha} = 10^\alpha$.

@@

Over the course of infection, mutation occurs within the host. 
However, we follow Shirreff \etal\ in assuming that SPVL of the strain transmitted by an infected 
individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation. 
Instead, the mutational effect is modeled as occurring in a single
step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.

We construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$. Finally, the probabilities are normalized so that each row sums to 1:
\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
\end{equation}
where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$. 
Unlike Shirreff \etal, who allowed for variation in the expressed phenotype (SPVL) of each genotype,
we use a one-to-one genotype-phenotype map. Thus there is a single value for transmission rate and
for progression rate corresponding to each SPVL compartment in the model:
\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i),\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
\end{aligned}
\end{equation}

@@

Lastly, we incorporate heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1). We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
For calibration purposes, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. 
(For the heterogeneous model, approximately
10\% of the samples failed due to numerical instability.)

@@

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory. Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer exponential-growth phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c). In this case the delay in the peak is more pronounced than for low $I(0)$, because the rate of SPVL increase is eventually limited by the mutation rate. The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase. However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.

Across the entire range of parameters covered by the LHS analysis, all
of the classes of models we considered produce qualitatively similar
virulence trajectories, which we quantify in terms of population mean
\Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean
\Lspvl\ corresponds to higher virulence). Although the speed of
virulence evolution varies, leading to wide variation in the peak SPVL
(ranging from 3 to 5.5) because HIV can evolve farther toward
its exponential-phase optimum before the transition to the
endemic phase, virulence peaks in all models between 200 and
300 years.

@@

One of the distinguishing features of the scaled parameters (in comparison to the unscaled parameters) is that the range of the parameters is compressed (relative to other models) for models without extra-pair contact as those models require parameters to be scaled by a large value in order to achieve a given $r$ of 0.04.
Models with extra-pair contact show wide range parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require wide range of scaling factors to achieve the target $r$.
Parameter range for random-mixing model is severely compressed because latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.

>>>>

\todo{Define virulence? I think it might be a good idea to add it to author summary as well for non-technical readers...}
The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
pathogens with higher reproduction numbers ($\rzero$) --- that is, the number of secondary infections caused by a single infected host over the course of its infectious period --- will tend to increase in prevalence relative to strains with lower reproductive ratios.
Pathogens can increase their reproduction numbers either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
transmission and disease-induced mortality rates are both driven by the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.

Fraser \etal\ have applied these ideas to HIV,
showing that the virus appears to satisfy the prerequisites of
the tradeoff theory.
The \emph{set-point viral load} (SPVL: i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection) is a measurable proxy for the rate of HIV within-host
reproduction. Higher viral loads are correlated with 
shorter progression times (higher virulence). \todo{We say the same thing in the next sentence.}
Studies of discordant partnerships --- stable
sexual partnerships with one infected and one uninfected partner --- %
have shown that SPVL is (1) correlated with faster transmission
(the time to infection of the uninfected partner is shorter when
their partner has a higher SPVL) and (2) heritable
(when the uninfected member of
the couple does become infected, their SPVL is similar to their
partner's).
Furthermore, the rate of transmission increases with
increasing SPVL (and hence with decreasing progression time)
but with a decreasing slope as progression time decreases,
fulfilling the requirements of the tradeoff theory
 \cite{Fraser+2007}.
Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

Our models
explicitly track the evolution of the distribution of $\log_{10}$ SPVL
(which we denote as $\alpha$)
rather than the rate of progression to
AIDS itself (hereafter we use ``virulence'' to denote \Lspvl).
We use a single-stage model of HIV that assumes constant
infectivity over the course of an exponentially
distributed infectious period. This assumption contrasts
with Shirreff \etal's previous model which
explicitly tracked three stages of HIV infection (primary,
asymptomatic, and AIDS) and used a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. 
Shirreff \etal\ allowed the transmission rate
and infection duration to depend on virus load only during the
asymptomatic stage. In order to adapt their parameterization to our
single-stage model, we used their estimates to compute the SPVL-dependent transmission
rate and duration during the asymptomatic stage and 
then derived the overall duration of the infectious period
as the sum of the three stage durations and the 
average transmission rate as the duration-weighted 
average of the three stage-specific transmission rates. 
Thus the within-couple transmission rate, $\beta$
(see ``Contact Structure'' below),
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL
(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

\todo{I changed $V_\alpha$ to $V(\alpha)$ because $\alpha$ isn't necessarily an integer...}

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V(\alpha) ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V(\alpha) ^ {\beta_k}}%
{V(\alpha)^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\label{eq:hillfuns}
\end{equation}
where $V(\alpha) = 10^\alpha$.

@@

Over the course of infection, mutation occurs within the host. 
However, we follow Shirreff \etal\ in assuming that SPVL of the strain transmitted by an infected 
individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation. 
Instead, the mutational effect is modeled as occurring in a single
step at the time of transmission.

First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.

We then construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$. Finally, the probabilities are normalized so that each row sums to 1:
\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
\end{equation}
where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$. 
Unlike Shirreff \etal, who allowed for variation in the expressed phenotype (SPVL) of each genotype,
we use a one-to-one genotype-phenotype map. Thus there is a single value for transmission rate and
for progression rate corresponding to each SPVL compartment in the model:
\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i),\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
\end{aligned}
\end{equation}

@@

Lastly, we incorporate heterogeneity in sexual activity into the models. Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1) \todo{We should also mention that appendix S1 has full model details... It doesn't seem like we mention that anywhere in the txet.}. We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}. 
(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity. As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 \todo{We mention that initial sd is 0.2 in the Appendix but we don't say it in the text. We can add it here or to the table...} and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
For calibration purposes, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. 
(For the heterogeneous model, approximately
10\% of the samples failed due to numerical instability.)

@@

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material) \todo{I'm not sure where in supplementary material you're referring to... Isn't this just figure 1?}, has little effect on the virulence trajectory. Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer exponential-growth phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c). In this case the delay in the peak is more pronounced than for low $I(0)$, because the rate of SPVL increase is eventually limited by the mutation rate. The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase. However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase. \todo{The last sentence makes sense but it's a little bit awkward... I'm not sure why...}

Across the entire range of parameters covered by the LHS analysis \todo{replace this with Latin hypercube samples? Seems like a better terminology than LHS analysis}, all
of the classes of models we considered produce qualitatively similar
virulence trajectories, which we quantify in terms of population mean
\Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean
\Lspvl\ corresponds to higher virulence). Although the speed of
virulence evolution varies, leading to wide variation in the peak SPVL \todo{log10 SPVL?}
(ranging from 3 to 5.5) because HIV can evolve farther toward
its exponential-phase optimum before the transition to the
endemic phase, virulence peaks in all models between 200 and
300 years.

@@

One of the distinguishing features of the scaled parameters (in comparison to the unscaled parameters) is that the range of the parameters is compressed (relative to other models) for models without extra-pair contact as those models require parameters to be scaled by a large value in order to achieve a given $r$ of 0.04.
Models with extra-pair contact show wide range parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require wide range of scaling factors to achieve the target $r$.
Parameter range for random-mixing model is severely compressed because Latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.
